The role of emotional factors in face processing abilities in Autism Spectrum Conditions

### 2023-07-17 manuscript.Rmd
Added Discussion

The Hospital Anxiety and Depression scale [HADS, @zigmond1983a] consists of seven questions for anxiety and seven questions for depression. Depression and anxiety scales are scored separately. The HADS has been shown to perform well in measuring anxiety and depression symptoms in non-clinical samples as well as clinical samples [@bjelland2002a], including the identification of anxiety and depression in autistic individuals, with good internal consistency and convergent validity for both the anxiety and depression scales [@uljarevic2013a]. The means, standard deviations and distribution of the HADS data for our data from typically developed adults confirm the measure is normally distributed, with suitable variation in scores that will allow for the exploration of the impacts of anxiety and depression on our variables of interest. For HADS anxiety, M = `r round(mean(TD$HANX7),digits=2)` (SD = `r round(sd(TD$HANX7),digits=2)`). For HADS depression, M = `r round(mean(TD$HDEP7),digits=2)` (SD = `r round(sd(TD$HDEP7),digits=2)`).

@@

To test the contribution of autistic traits, alexithymic traits and mood disorder symptomatology to predicting emotion and identity recognition, we conducted a series of hierarchical regressions. These are summarised in Figure 1.

@@

powerL <- pwr.f2.test(u = 6, f2=0.15,sig.level=0.05,power=0.9)
N <- 1 + powerL$u + ceiling(powerL$v)

powerR <- pwr.r.test(r=0.383,sig.level=0.05,power=0.9,alternative='two.sided')

@@

A priori analyses were carried out to establish the required sample size for a regression analysis, which will be our core hypothesis-testing statistical procedure. The power analysis was carried out using the software G*Power [@faul2009a], with 1 number of tested predictors and 6 total predictors, and an effect size ($f^2$) was 0.15. This power analyses thus tests the power for a predictor to reveal a significant independent contribution to a model, over and above previous entered predictors. Results showed that a sample size of 73 participants would be needed for a significance level of 0.05 and the level of power at 0.90. Also, power analyses states that a sample size of `r ceiling(powerR$n)` would be required to establish a correlation between emotion recognition and the TAS-20 (the alexithymia measure).

@@

@@

Data associated with the non-autistic sample is available on the Open Science Framework. This is the link: [https://osf.io/rhykm/?view_only=5585f03089764c9f98427b7b012c9bc0](https://osf.io/rhykm/?view_only=5585f03089764c9f98427b7b012c9bc0)

>>>>

The Hospital Anxiety and Depression scale [HADS, @zigmond1983a] consists of seven questions for anxiety and seven questions for depression. Depression and anxiety scales are scored separately. The HADS has been shown to perform well in measuring anxiety and depression symptoms in non-clinical samples as well as clinical samples [@bjelland2002a], including the identification of anxiety and depression in autistic individuals, with good internal consistency and convergent validity for both the anxiety and depression scales [@uljarevic2013a]. The means, standard deviations and distribution of the HADS data for our data from typically developed adults confirm the measure is normally distributed, with suitable variation in scores that will allow for the exploration of the impacts of anxiety and depression on our variables of interest.

@@

To test the contribution of autistic traits, alexithymic traits and mood disorder symptomatology to predicting emotion and identity recognition, we conducted a series of hierarchical regressions. These are summarised in Figure \@ref(fig:figure1).

```{r figure1, out.width="80%",, fig.cap="Diagram showing the hierarchical regression series analyses. ASD-SI = self-identifying autistics, ASD-CD = clinically diagnosed autistics, NA = non-autistics,  AQ-28 = Autism Quotient, TAS-20 = Toronto Alexithymia Scale, HADS = Hospital Anxiety and Depression Scale, PSE = Point of subjective equivalence, ATT = Attribution Threshold.", fig.align="center", include=TRUE, echo=FALSE}

knitr::include_graphics('Fig1.jpg')

@@

powerL <- pwr.f2.test(u = 1, f2=0.15,sig.level=0.05,power=0.9)
N <- 1 + powerL$u + ceiling(powerL$v)

powerR <- pwr.r.test(r=0.3,sig.level=0.05,power=0.8,alternative='greater')

@@

A priori analyses were carried out to establish the required sample size for a regression analysis, which will be our core hypothesis-testing statistical procedure. The power analysis was carried out using the software G*Power [@faul2009a], with 1 number of tested predictors and 6 total predictors, and an effect size ($f^2$) was `r powerL$f2`. This power analyses thus tests the power for a predictor to reveal a significant independent contribution to a model, over and above previous entered predictors. Results showed that a sample size of `r N` participants would be needed for a significance level of `r powerL$sig.level` and the level of power at `r powerL$power`. Also, power analyses states that a sample size of `r ceiling(powerR$n)` would be required to establish a correlation between emotion recognition and the TAS-20 (the alexithymia measure).

@@

This study aimed to investigate the extent to which co-occurring alexithymia and/or mood disorder symptomatology predict atypical facial emotion recognition in autistic and non-autistic individuals. It was predicted that alexithymia would be associated with poorer emotion recognition (rather than autism). We also predicted that anxiety and depression may also contribute to differences seen in emotion recognition abilities between autistic and non-autistic individuals. While we did find the expected correlation between alexithymia and autistic traits, neither alexithymia nor autism (or autistic traits) contributed significantly to emotion or identity recognition abilities in any of the analyses conducted (including analyses using newly proposed alexithymia measures, with outliers included, etc). This was also true for anxiety and depression. The only significant predictor was gender: gender was a significant predictor for emotion PSE scores, and follow-up comparisons indicated that the ‘other’ gender group had significantly lower emotion PSE scores compared to the male and female groups.

While autistic traits did show a small zero order correlation with emotion attribution scores, this association did not survive controlling for demographic variables in our regression analyses. Autistic group membership also did not predict face processing abilities in our regression analyses. Previous literature has documented emotion processing difficulties in autism [@uljarevic2013a], but noted that studies are far from consistent: not all autistic individuals experience differences or impairments in their socio-emotional abilities, reflecting the heterogeneity of the condition [@baron-cohen1997a]. Indeed, one motivation for the present study was to consider the argument that previously reported effects of autism were actually driven by alexithymia [@cook2013a]. However, no effect on, or association with, our face processing task was found with alexithymia either.

Plausibly, the relationship between alexithymia and face processing may be more complex than a simple linear effect: some research into alexithymia and other socio-emotional abilities indicates that the role of alexithymia may be more mediatory, or may not influence atypical socio-emotional abilities at all [@brett2022a; @morie2019a; @shah2019a]. Furthermore, other authors have posited that proposed explanatory power of alexithymia in socioemotional problems in autism have been overstated [@shah2019a]. While alexithymia and emotion recognition abilities have been the topic of numerous studies, with both clinical and non-clinical populations, to the best of our knowledge there has yet to be a meta-analysis that has established the current evidence base on the relationship between alexithymia and emotion recognition, and, importantly, assessed the degree to which publication bias may be present in this literature. This need was discussed in a systematic review of the role of alexithymia in emotional face processing [@grynberg2012a], but this research need appears to have been as yet unmet. It is a particular strength of the registered report format that studies are assessed and accepted in principle prior to data collection and before results are known: we contemplate that there may be other unpublished studies similar to ours that failed to find effects of alexithymia on emotion processing.

Mood disorder symptomatology did not predict emotion processing either. Previous research with non-autistic individuals indicated that anxiety and depression may affect emotion recognition abilities [@charidza2022a; @dalili2015a], and so with the relatively high prevalence rates of both anxiety and depression in autism, it was thought that mood disorders would contribute to some degree in this study. However, other studies have indicated that females with and without depression perform similarly on facial emotion recognition tasks [@fieker2016a]. Some research indicates that anxiety has little effect on emotion recognition in autism: Wong et al. [-@wong2012a] investigated social anxiety and emotion recognition in a sample of autistic children and found that autistic children performed poorer than non-autistic controls on an emotion recognition task, but that anxiety was not a contributing factor. Plausibly, our measures of depression and anxiety may have lacked sufficiently specificity to uncover more nuanced relationships: for example, social worries and fear of negative evaluation has been shown to impact eye gaze behaviour in autism [@white2012a]. Of course, given that we did not find strong effects of autism in face processing abilities, there is not an autism effect for anxiety or depression to explain in the current data.

One factor that did predict some of our face processing measures was gender. The effect of gender was not predicted, although previous research into alexithymia, autistic traits and empathy did highlight that gender may have an important predictive role in empathy abilities [@shah2019a]. Possibly, effects of gender diversity could be to do with neurodivergence. An association between gender diversity and neurodiversity has been observed [@vlerken2020a], and in the present study, we did note that the “other” gender group had the highest autistic traits of all three gender groups, and our autistic sample had twice as many people who considered themselves to have an other gender, compared to the non-autistic group. The measure on which gender was predictive was emotion PSE: specifically, it indicated that those who described their gender as other perceived the morphed faces to be angry earlier in the morph spectrum than males or females, indicating a bias (relative to males and females) to see anger rather than disgust. While interesting, we would emphasise that this does not mean that face processing is deficient in those who do not identify as male or female, and, rather than necessarily due to associations with neurodivergence, could be a product of these individuals’ social experiences. We could speculate that individuals who do not identify as male or female may be more likely to have been in threatening situations, been the subject of bullying or violence: such experiences could be expected to lead to a heightened sensitivity to anger in others. Indeed, transgender people are at an increased likelihood of being victims of multiple episodes of violence [@messinger2022a]. However, our gender category of “other” is very vague, and limits our ability to draw form conclusions between individuals’ gender identities and emotion processing abilities.

Overall, the implications of the current study indicate that neither alexithymia, autism, nor mood disorders significantly impacted the facial emotion recognition abilities of autistic individuals. While this differs to previous research, and opens up avenues for further examination, it also begs the question whether there are other factors contributing to atypical facial emotion recognition abilities. Gender appeared to have an effect on emotion recognition abilities in this sample, specifically individuals who identified as neither male nor female, and so future research could focus on gender diversity, autism, and social-emotional cognition. Given the expansive research that has taken place on alexithymia and emotion processing abilities, we argue that a meta-analysis and assessment of previous publication bias would be timely.

One consideration is that this study was conducted online. While the effectiveness of collecting data online and the quality of the data has been questioned, it appears that there is little to no statistical difference between collecting data online or in-person [@sauter2022a; @uittenhove2022a]. Therefore, it is unlikely that using an online design would have reduced the quality of the current experiment. However, our online approach does raise two issues for reflection.

Firstly, unfortunately in our recruitment of the autistic sample, it came apparent that there were fraudulent participants attempting to take part multiple times. Thankfully, this was noticed and none of this sample were used in the final data set. While the current study was able to rectify issues with fraudulent participants, it does appear that issue is happening increasingly with research. @pellicano2023a was also affected by ‘scammer participants’ pretending to be autistic individuals, or parents of autistic individuals, during online interviews. We echo the warnings of @pellicano2023a that autism researchers need to be alert to the real threat of imposter participants, especially in online research.

Secondly, it is plausible that effects of autism and alexithymia seen in previous face-to-face research may not have been observed in our online study because of delivering the experiment in this modality. Plausibly, the ability to participate from their own homes in a more comfortable environment relative to most research laboratories, to self-pace (to some extent), and without perceptions of judgement from any research staff, autistic individuals may perform quite well on emotion processing tasks such as ours. @sucksmith2013a asked autistic adults without intellectual disability to complete an online facial emotion task, and suggested that the participants may have felt more comfortable at home compared to a lab environment, and that this increases the validity of the results, as it is more reflective of the participants natural abilities. We concur that being able to take part at home, away from the lab, may have reduced apparent deficits in face processing, although how “valid” a reflection of ability depends on under what circumstances individuals are usually processing faces in their daily lives. Certainly understanding the role of general environment factors and experiment modality is important for the field to connect research findings to real world difficulties, as online research may diverge from participant reports or clinical observations of finding face processing hard “in real life”.

As part of conducting our study online, we opted in to include individuals who self-identify as autistic, given the practical barriers verifying formal diagnoses even if we insisted on all participants having a clinical diagnosis before taking part. There was no “penalty” to participants if they reported that they self-identified as autistic, so we consider it unlikely that participants would have been motivated to be dishonest about their autism diagnostic status. None of our analyses suggested that including self-identifying individuals reduced our effects: indeed, in one analysis where we removed autistic adults who did not score above the AQ threshold, the participants who did not meet this threshold all reported to hold formal diagnoses. Broadly our experience would support the future inclusion of self-identifying autistic adults in research studies, particularly as this may make research more inclusive for subgroups who struggle to attain a formal diagnosis (e.g. women).

One limitation of the present study is that the non-autistic group showed surprisingly high levels of autistic and alexithymic traits. 46.9\% and 23.5\% of our “non-autistic” sample scored above threshold on the AQ and TAS-20 respectively (by comparison, 97.5\% and 52.5\% of the autistic group scored above threshold on these measures, which is in line with previous reports; @hoekstra2011a; @kinnaird2019a). While it suggests our non-autism sample may have been unusual in some way, our regression-based analyses should still have been able to uncover relationships between alexithymia, autistic traits and our face processing measures as there should still have been sufficient variation at the lower end of these scales for associations to be able to show themselves. Thus, we do not believe atypicalities in the non-autistic sample can account for the lack of effects.

In sum, our present study did not find the expected effects of alexithymia on emotion face recognition. Weak associations between autistic traits and emotion recognition did not survive controlling for other variables in our analyses. This adds to a growing picture of heterogenous results regarding autism and emotion recognition, but does not support alexithymia as a strong candidate for explaining this heterogeneity. Future meta-analyses and a review of the evidence base for a relationship between alexithymia and emotion recognition is warranted. We also did not observe associations with mood disorder symptomatology, and posit that future research may need more specific measures of specific aspects of mood disorders such as social anxiety to evaluate the role of different forms of mood disorders in emotion recognition. Future research should focus on establishing the influence of gender diversity on socio-emotional abilities in autistic spectrum conditions though we speculate that social experience rather than neurodiversity per se may explain why gender non-conforming individuals showed differences in their perceptions of anger and disgust.

@@

Our data and analysis code are available to view on the OSF:

### 2023-07-17 manuscript.tex
Added Discussion

The Hospital Anxiety and Depression scale (HADS, \protect\hyperlink{ref-zigmond1983a}{Zigmond and Snaith, 1983}) consists of seven questions for anxiety and seven questions for depression. Depression and anxiety scales are scored separately. The HADS has been shown to perform well in measuring anxiety and depression symptoms in non-clinical samples as well as clinical samples (\protect\hyperlink{ref-bjelland2002a}{Bjelland et al., 2002}), including the identification of anxiety and depression in autistic individuals, with good internal consistency and convergent validity for both the anxiety and depression scales (\protect\hyperlink{ref-uljarevic2013a}{Uljarevic and Hamilton, 2013}). The means, standard deviations and distribution of the HADS data for our data from typically developed adults confirm the measure is normally distributed, with suitable variation in scores that will allow for the exploration of the impacts of anxiety and depression on our variables of interest. For HADS anxiety, M = 10.77 (SD = 2.93). For HADS depression, M = 8.68 (SD = 2.06).

@@

To test the contribution of autistic traits, alexithymic traits and mood disorder symptomatology to predicting emotion and identity recognition, we conducted a series of hierarchical regressions. These are summarised in Figure 1.

@@

A priori analyses were carried out to establish the required sample size for a regression analysis, which will be our core hypothesis-testing statistical procedure. The power analysis was carried out using the software G*Power (\protect\hyperlink{ref-faul2009a}{Faul et al., 2009}), with 1 number of tested predictors and 6 total predictors, and an effect size (\(f^2\)) was 0.15. This power analyses thus tests the power for a predictor to reveal a significant independent contribution to a model, over and above previous entered predictors. Results showed that a sample size of 73 participants would be needed for a significance level of 0.05 and the level of power at 0.90. Also, power analyses states that a sample size of 67 would be required to establish a correlation between emotion recognition and the TAS-20 (the alexithymia measure).

@@

@@

Data associated with the non-autistic sample is available on the Open Science Framework. This is the link: \url{https://osf.io/rhykm/?view_only=5585f03089764c9f98427b7b012c9bc0}

@@

@@

@@

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-gould2011a}{}}%
Gould J, Ashton-Smith J. 2011. Missed diagnosis or misdiagnosis? Girls and women on the autism spectrum.

@@

\leavevmode\vadjust pre{\hypertarget{ref-ko2018a}{}}%
Ko B. 2018. A brief review of facial emotion recognition based on visual information. \emph{Sensors} \textbf{18}:401.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-murphy2016a}{}}%
Murphy CM, Wilson CE, Robertson DM, Ecker C, Daly EM, Hammond N, McAlonan GM. 2016. Autism spectrum disorder in adults: Diagnosis, management, and health services development. Neuropsychiatric disease and treatment.

@@

\leavevmode\vadjust pre{\hypertarget{ref-oakley2020a}{}}%
Oakley BF, Jones EJ, Crawley D, Charman T, Buitelaar J, Tillmann J, Loth E. 2020. Alexithymia in autism: Cross-sectional and longitudinal associations with social-communication difficulties, anxiety and depression symptoms. \emph{Psychological Medicine} 1--13.

@@

@@

@@

@@

@@

@@

@@

>>>>

The Hospital Anxiety and Depression scale (HADS, \protect\hyperlink{ref-zigmond1983a}{Zigmond and Snaith, 1983}) consists of seven questions for anxiety and seven questions for depression. Depression and anxiety scales are scored separately. The HADS has been shown to perform well in measuring anxiety and depression symptoms in non-clinical samples as well as clinical samples (\protect\hyperlink{ref-bjelland2002a}{Bjelland et al., 2002}), including the identification of anxiety and depression in autistic individuals, with good internal consistency and convergent validity for both the anxiety and depression scales (\protect\hyperlink{ref-uljarevic2013a}{Uljarevic and Hamilton, 2013}). The means, standard deviations and distribution of the HADS data for our data from typically developed adults confirm the measure is normally distributed, with suitable variation in scores that will allow for the exploration of the impacts of anxiety and depression on our variables of interest.

@@

To test the contribution of autistic traits, alexithymic traits and mood disorder symptomatology to predicting emotion and identity recognition, we conducted a series of hierarchical regressions. These are summarised in Figure \ref{fig:figure1}.

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{Fig1}

}

\caption{Diagram showing the hierarchical regression series analyses. ASD-SI = self-identifying autistics, ASD-CD = clinically diagnosed autistics, NA = non-autistics,  AQ-28 = Autism Quotient, TAS-20 = Toronto Alexithymia Scale, HADS = Hospital Anxiety and Depression Scale, PSE = Point of subjective equivalence, ATT = Attribution Threshold.}\label{fig:figure1}
\end{figure}

@@

A priori analyses were carried out to establish the required sample size for a regression analysis, which will be our core hypothesis-testing statistical procedure. The power analysis was carried out using the software G*Power (\protect\hyperlink{ref-faul2009a}{Faul et al., 2009}), with 1 number of tested predictors and 6 total predictors, and an effect size (\(f^2\)) was 0.15. This power analyses thus tests the power for a predictor to reveal a significant independent contribution to a model, over and above previous entered predictors. Results showed that a sample size of 73 participants would be needed for a significance level of 0.05 and the level of power at 0.9. Also, power analyses states that a sample size of 67 would be required to establish a correlation between emotion recognition and the TAS-20 (the alexithymia measure).

@@

This study aimed to investigate the extent to which co-occurring alexithymia and/or mood disorder symptomatology predict atypical facial emotion recognition in autistic and non-autistic individuals. It was predicted that alexithymia would be associated with poorer emotion recognition (rather than autism). We also predicted that anxiety and depression may also contribute to differences seen in emotion recognition abilities between autistic and non-autistic individuals. While we did find the expected correlation between alexithymia and autistic traits, neither alexithymia nor autism (or autistic traits) contributed significantly to emotion or identity recognition abilities in any of the analyses conducted (including analyses using newly proposed alexithymia measures, with outliers included, etc). This was also true for anxiety and depression. The only significant predictor was gender: gender was a significant predictor for emotion PSE scores, and follow-up comparisons indicated that the `other' gender group had significantly lower emotion PSE scores compared to the male and female groups.

While autistic traits did show a small zero order correlation with emotion attribution scores, this association did not survive controlling for demographic variables in our regression analyses. Autistic group membership also did not predict face processing abilities in our regression analyses. Previous literature has documented emotion processing difficulties in autism (\protect\hyperlink{ref-uljarevic2013a}{Uljarevic and Hamilton, 2013}), but noted that studies are far from consistent: not all autistic individuals experience differences or impairments in their socio-emotional abilities, reflecting the heterogeneity of the condition (\protect\hyperlink{ref-baron-cohen1997a}{Baron-Cohen et al., 1997}). Indeed, one motivation for the present study was to consider the argument that previously reported effects of autism were actually driven by alexithymia (\protect\hyperlink{ref-cook2013a}{Cook et al., 2013}). However, no effect on, or association with, our face processing task was found with alexithymia either.

Plausibly, the relationship between alexithymia and face processing may be more complex than a simple linear effect: some research into alexithymia and other socio-emotional abilities indicates that the role of alexithymia may be more mediatory, or may not influence atypical socio-emotional abilities at all (\protect\hyperlink{ref-brett2022a}{Brett and Maybery, 2022}; \protect\hyperlink{ref-morie2019a}{Morie et al., 2019}; \protect\hyperlink{ref-shah2019a}{Shah et al., 2019}). Furthermore, other authors have posited that proposed explanatory power of alexithymia in socioemotional problems in autism have been overstated (\protect\hyperlink{ref-shah2019a}{Shah et al., 2019}). While alexithymia and emotion recognition abilities have been the topic of numerous studies, with both clinical and non-clinical populations, to the best of our knowledge there has yet to be a meta-analysis that has established the current evidence base on the relationship between alexithymia and emotion recognition, and, importantly, assessed the degree to which publication bias may be present in this literature. This need was discussed in a systematic review of the role of alexithymia in emotional face processing (\protect\hyperlink{ref-grynberg2012a}{Grynberg et al., 2012}), but this research need appears to have been as yet unmet. It is a particular strength of the registered report format that studies are assessed and accepted in principle prior to data collection and before results are known: we contemplate that there may be other unpublished studies similar to ours that failed to find effects of alexithymia on emotion processing.

Mood disorder symptomatology did not predict emotion processing either. Previous research with non-autistic individuals indicated that anxiety and depression may affect emotion recognition abilities (\protect\hyperlink{ref-charidza2022a}{Charidza and Gillmeister, 2022}; \protect\hyperlink{ref-dalili2015a}{Dalili et al., 2015}), and so with the relatively high prevalence rates of both anxiety and depression in autism, it was thought that mood disorders would contribute to some degree in this study. However, other studies have indicated that females with and without depression perform similarly on facial emotion recognition tasks (\protect\hyperlink{ref-fieker2016a}{Fieker et al., 2016}). Some research indicates that anxiety has little effect on emotion recognition in autism: Wong et al. (\protect\hyperlink{ref-wong2012a}{2012}) investigated social anxiety and emotion recognition in a sample of autistic children and found that autistic children performed poorer than non-autistic controls on an emotion recognition task, but that anxiety was not a contributing factor. Plausibly, our measures of depression and anxiety may have lacked sufficiently specificity to uncover more nuanced relationships: for example, social worries and fear of negative evaluation has been shown to impact eye gaze behaviour in autism (\protect\hyperlink{ref-white2012a}{White et al., 2012}). Of course, given that we did not find strong effects of autism in face processing abilities, there is not an autism effect for anxiety or depression to explain in the current data.

One factor that did predict some of our face processing measures was gender. The effect of gender was not predicted, although previous research into alexithymia, autistic traits and empathy did highlight that gender may have an important predictive role in empathy abilities (\protect\hyperlink{ref-shah2019a}{Shah et al., 2019}). Possibly, effects of gender diversity could be to do with neurodivergence. An association between gender diversity and neurodiversity has been observed (\protect\hyperlink{ref-vlerken2020a}{Vlerken et al., 2020}), and in the present study, we did note that the ``other'' gender group had the highest autistic traits of all three gender groups, and our autistic sample had twice as many people who considered themselves to have an other gender, compared to the non-autistic group. The measure on which gender was predictive was emotion PSE: specifically, it indicated that those who described their gender as other perceived the morphed faces to be angry earlier in the morph spectrum than males or females, indicating a bias (relative to males and females) to see anger rather than disgust. While interesting, we would emphasise that this does not mean that face processing is deficient in those who do not identify as male or female, and, rather than necessarily due to associations with neurodivergence, could be a product of these individuals' social experiences. We could speculate that individuals who do not identify as male or female may be more likely to have been in threatening situations, been the subject of bullying or violence: such experiences could be expected to lead to a heightened sensitivity to anger in others. Indeed, transgender people are at an increased likelihood of being victims of multiple episodes of violence (\protect\hyperlink{ref-messinger2022a}{Messinger et al., 2022}). However, our gender category of ``other'' is very vague, and limits our ability to draw form conclusions between individuals' gender identities and emotion processing abilities.

Overall, the implications of the current study indicate that neither alexithymia, autism, nor mood disorders significantly impacted the facial emotion recognition abilities of autistic individuals. While this differs to previous research, and opens up avenues for further examination, it also begs the question whether there are other factors contributing to atypical facial emotion recognition abilities. Gender appeared to have an effect on emotion recognition abilities in this sample, specifically individuals who identified as neither male nor female, and so future research could focus on gender diversity, autism, and social-emotional cognition. Given the expansive research that has taken place on alexithymia and emotion processing abilities, we argue that a meta-analysis and assessment of previous publication bias would be timely.

One consideration is that this study was conducted online. While the effectiveness of collecting data online and the quality of the data has been questioned, it appears that there is little to no statistical difference between collecting data online or in-person (\protect\hyperlink{ref-sauter2022a}{Sauter et al., 2022}; \protect\hyperlink{ref-uittenhove2022a}{Uittenhove et al., 2022}). Therefore, it is unlikely that using an online design would have reduced the quality of the current experiment. However, our online approach does raise two issues for reflection.

Firstly, unfortunately in our recruitment of the autistic sample, it came apparent that there were fraudulent participants attempting to take part multiple times. Thankfully, this was noticed and none of this sample were used in the final data set. While the current study was able to rectify issues with fraudulent participants, it does appear that issue is happening increasingly with research. Pellicano et al. (\protect\hyperlink{ref-pellicano2023a}{2023}) was also affected by `scammer participants' pretending to be autistic individuals, or parents of autistic individuals, during online interviews. We echo the warnings of Pellicano et al. (\protect\hyperlink{ref-pellicano2023a}{2023}) that autism researchers need to be alert to the real threat of imposter participants, especially in online research.

Secondly, it is plausible that effects of autism and alexithymia seen in previous face-to-face research may not have been observed in our online study because of delivering the experiment in this modality. Plausibly, the ability to participate from their own homes in a more comfortable environment relative to most research laboratories, to self-pace (to some extent), and without perceptions of judgement from any research staff, autistic individuals may perform quite well on emotion processing tasks such as ours. Sucksmith et al. (\protect\hyperlink{ref-sucksmith2013a}{2013}) asked autistic adults without intellectual disability to complete an online facial emotion task, and suggested that the participants may have felt more comfortable at home compared to a lab environment, and that this increases the validity of the results, as it is more reflective of the participants natural abilities. We concur that being able to take part at home, away from the lab, may have reduced apparent deficits in face processing, although how ``valid'' a reflection of ability depends on under what circumstances individuals are usually processing faces in their daily lives. Certainly understanding the role of general environment factors and experiment modality is important for the field to connect research findings to real world difficulties, as online research may diverge from participant reports or clinical observations of finding face processing hard ``in real life''.

As part of conducting our study online, we opted in to include individuals who self-identify as autistic, given the practical barriers verifying formal diagnoses even if we insisted on all participants having a clinical diagnosis before taking part. There was no ``penalty'' to participants if they reported that they self-identified as autistic, so we consider it unlikely that participants would have been motivated to be dishonest about their autism diagnostic status. None of our analyses suggested that including self-identifying individuals reduced our effects: indeed, in one analysis where we removed autistic adults who did not score above the AQ threshold, the participants who did not meet this threshold all reported to hold formal diagnoses. Broadly our experience would support the future inclusion of self-identifying autistic adults in research studies, particularly as this may make research more inclusive for subgroups who struggle to attain a formal diagnosis (e.g.~women).

One limitation of the present study is that the non-autistic group showed surprisingly high levels of autistic and alexithymic traits. 46.9\% and 23.5\% of our ``non-autistic'' sample scored above threshold on the AQ and TAS-20 respectively (by comparison, 97.5\% and 52.5\% of the autistic group scored above threshold on these measures, which is in line with previous reports; Hoekstra et al. (\protect\hyperlink{ref-hoekstra2011a}{2011}); Kinnaird et al. (\protect\hyperlink{ref-kinnaird2019a}{2019})). While it suggests our non-autism sample may have been unusual in some way, our regression-based analyses should still have been able to uncover relationships between alexithymia, autistic traits and our face processing measures as there should still have been sufficient variation at the lower end of these scales for associations to be able to show themselves. Thus, we do not believe atypicalities in the non-autistic sample can account for the lack of effects.

In sum, our present study did not find the expected effects of alexithymia on emotion face recognition. Weak associations between autistic traits and emotion recognition did not survive controlling for other variables in our analyses. This adds to a growing picture of heterogenous results regarding autism and emotion recognition, but does not support alexithymia as a strong candidate for explaining this heterogeneity. Future meta-analyses and a review of the evidence base for a relationship between alexithymia and emotion recognition is warranted. We also did not observe associations with mood disorder symptomatology, and posit that future research may need more specific measures of specific aspects of mood disorders such as social anxiety to evaluate the role of different forms of mood disorders in emotion recognition. Future research should focus on establishing the influence of gender diversity on socio-emotional abilities in autistic spectrum conditions though we speculate that social experience rather than neurodiversity per se may explain why gender non-conforming individuals showed differences in their perceptions of anger and disgust.

@@

Our data and analysis code are available to view on the OSF:

@@

\leavevmode\vadjust pre{\hypertarget{ref-brett2022a}{}}%
Brett JD, Maybery MT. 2022. Understanding oneself to understand others: The role of alexithymia and anxiety in the relationships between autistic trait dimensions and empathy. \emph{Journal of Autism and Developmental Disorders} \textbf{52}:1971--1983.

@@

\leavevmode\vadjust pre{\hypertarget{ref-charidza2022a}{}}%
Charidza CA, Gillmeister H. 2022. Differential beta desynchronisation responses to dynamic emotional facial expressions are attenuated in higher trait anxiety and autism. \emph{Cognitive, Affective, \& Behavioral Neuroscience} \textbf{22}:1404--1420.

@@

\leavevmode\vadjust pre{\hypertarget{ref-dalili2015a}{}}%
Dalili MN, Penton-Voak IS, Harmer CJ, Munafò MR. 2015. Meta-analysis of emotion recognition deficits in major depressive disorder. \emph{Psychological medicine} \textbf{45}:1135--1144.

@@

\leavevmode\vadjust pre{\hypertarget{ref-fieker2016a}{}}%
Fieker M, Moritz S, Köther U, Jelinek L. 2016. Emotion recognition in depression: An investigation of performance and response confidence in adult female patients with depression. \emph{Psychiatry research} \textbf{242}:226--232.

@@

\leavevmode\vadjust pre{\hypertarget{ref-gould2011a}{}}%
Gould J, Ashton-Smith J. 2011. Missed diagnosis or misdiagnosis? Girls and women on the autism spectrum. \emph{Good Autism Practice (GAP} \textbf{12}:34--41.

\leavevmode\vadjust pre{\hypertarget{ref-grynberg2012a}{}}%
Grynberg D, Chang B, Corneille O, Maurage P, Vermeulen N, Berthoz S, Luminet O. 2012. Alexithymia and the processing of emotional facial expressions (EFEs): Systematic review, unanswered questions and further perspectives.

@@

\leavevmode\vadjust pre{\hypertarget{ref-ko2018a}{}}%
Ko B. 2018. A brief review of facial emotion recognition based on visual information. \emph{sensors} \textbf{18}:401.

@@

\leavevmode\vadjust pre{\hypertarget{ref-messinger2022a}{}}%
Messinger AM, Guadalupe-Diaz XL, Kurdyla V. 2022. Transgender polyvictimization in the US transgender survey. \emph{Journal of interpersonal violence} \textbf{37}:18810--18836.

@@

\leavevmode\vadjust pre{\hypertarget{ref-murphy2016a}{}}%
Murphy CM, Wilson CE, Robertson DM, Ecker C, Daly EM, Hammond N, McAlonan GM. 2016. Autism spectrum disorder in adults: Diagnosis, management, and health services development. \emph{Neuropsychiatric disease and treatment}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-oakley2020a}{}}%
Oakley BF, Jones EJ, Crawley D, Charman T, Buitelaar J, Tillmann J, Loth E. 2020. Alexithymia in autism: Cross-sectional and longitudinal associations with social-communication difficulties, anxiety and depression symptoms. Psychological Medicine.

@@

\leavevmode\vadjust pre{\hypertarget{ref-pellicano2023a}{}}%
Pellicano E, Adams D, Crane L, Hollingue C, Allen C, Almendinger K, Wheeley E. 2023. A possible threat to data integrity for online qualitative autism research. \emph{Autism} \textbf{13623613231174543}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-sauter2022a}{}}%
Sauter M, Stefani M, Mack W. 2022. Equal quality for online and lab data: A direct comparison from two dual-task paradigms. \emph{Open Psychology} \textbf{4}:47--59. doi:\href{https://doi.org/10.1515/psych-2022-0003}{10.1515/psych-2022-0003}

@@

\leavevmode\vadjust pre{\hypertarget{ref-sucksmith2013a}{}}%
Sucksmith E, Allison C, Baron-Cohen S, Chakrabarti B, Hoekstra RA. 2013. Empathy and emotion recognition in people with autism, first-degree relatives, and controls. \emph{Neuropsychologia} \textbf{51}:98--105.

@@

\leavevmode\vadjust pre{\hypertarget{ref-uittenhove2022a}{}}%
Uittenhove K, Jeanneret S, Vergauwe E. 2022. From lab-based to web-based behavioural research: Who you test is more important than how you test. \emph{PsyArXiv, February} \textbf{16}. doi:\href{https://doi.org/10.31234/osf.io/uy4kb}{10.31234/osf.io/uy4kb}

@@

\leavevmode\vadjust pre{\hypertarget{ref-vlerken2020a}{}}%
Vlerken RH, Fuchs CE, Miesen AI. 2020. Caring for transgender and gender diverse youth with co-occurring neurodiversityPediatric Gender Identity: Gender-Affirming Care for Transgender \& Gender Diverse Youth. pp. 137--148.

@@

\leavevmode\vadjust pre{\hypertarget{ref-white2012a}{}}%
White SW, Schry AR, Maddox BB. 2012. Brief report: The assessment of anxiety in high-functioning adolescents with autism spectrum disorder. \emph{Journal of Autism and Developmental Disorders} \textbf{42}:1138--1145.

@@

\leavevmode\vadjust pre{\hypertarget{ref-wong2012a}{}}%
Wong N, Beidel DC, Sarver DE, Sims V. 2012. Facial emotion recognition in children with high functioning autism and children with social phobia. \emph{Child Psychiatry \& Human Development} \textbf{43}:775--794.
